---
title: "Climate Exercise"
author: "Carl Boettiger & Dana Seidel"
output: github_document
---



# Unit I: Climate Change Module

## Warm-up exercise: Examining CO2 trends in R

- Example from <http://climate.nasa.gov/vital-signs/carbon-dioxide/>
- Raw data from <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt>

```{r message=FALSE}
library(tidyverse)
library(dplyr)
```

```{r message=FALSE}

co2 <- 
read_table("co2_mm_mlo.txt", 
                  comment="#",
                  col_names = c("year", "month", "decimal_date", "average",
                                "interpolated", "trend", "days"),
                  na = c("-1", "-99.99"))
co2
```


```{r}
ggplot(co2, aes(x = decimal_date, y = average)) + geom_line() 

```

Which months are the CO2 values at the maximum? Minimum?  Why is this?

The month with maximum CO2 is May, and the months with minimum CO2 is September and October. The reason is that the plants will grow in May when they will remove much CO2 from the atmosphere. However, in the fall, the plants cease growing and the CO2 concentrations increase again. 

What rolling average is used in computing the "trend" line?  How does the trend depend on the rolling average?

The yearly rolling average data is used in computing the "trend" line. 

------------------------

# Exercise I: Temperature Data

Each of the last years has consecutively set new records on global climate.  In this section we will analyze global mean temperature data.

Data from: <http://climate.nasa.gov/vital-signs/global-temperature>

## Question 1:

Describe the data set to the best of your ability given the documentation provided.  Describe what kind of column each data contains and what units it is measured in.  Then address our three key questions in understanding this data:

The data set shows the difference between annual average temperature and the 30-year average temperature of 1951-1980 in the period of 1880 to 2020. The data set contains a column of raw temperature difference and a column of temperature difference obtained from smoothing. The unit is celsius degree. 

- How are the measurements made? What is the associated measurement uncertainty?

The measurements are made by subtracting the annual average temperature from the 1951-1980 average temperature. An uncertainty is the accuracy of the measuring instruments as they must have changed in such a long period.

- What is the resolution of the data?

Each data point keeps two decimal places.

- Are their missing values? How should they be handled?

There is no missing value. If there is, it can be handled by assigning it the average value of the five or ten data points surrounding it.

## Question 2:

Construct the necessary R code to import and prepare for manipulation the following data set: <http://climate.nasa.gov/system/internal_resources/details/original/647_Global_Temperature_Data_File.txt>

```{r}
gt <- read_table("647_Global_Temperature_Data_File.txt", 
                  comment="#",
                 col_names = c("year","NoSmoothing","Lowess"))
gt
```

## Question 3:

Plot the trend in global mean temperatures over time.  Describe what you see in the plot and how you interpret the patterns you observe.

```{r}
ggplot(gt, aes(x =year, y =Lowess)) + geom_line() 
```
The trend line is smoother than the average temperature line. Since each data entry of the temporal column (trend) is closer to each other than the column of month, the plotted line will then be smoother with less fluctuations. The line is to demonstrate an overall rising pattern of global temperature regardless of the fluctuations in each year.

## Question 4: Evaluating the evidence for a "Pause" in warming?

The [2013 IPCC Report](https://www.ipcc.ch/pdf/assessment-report/ar5/wg1/WG1AR5_SummaryVolume_FINAL.pdf) included a tentative observation of a "much smaller increasing trend" in global mean temperatures since 1998 than was observed previously.  This led to much discussion in the media about the existence of a "Pause" or "Hiatus" in global warming rates, as well as much research looking into where the extra heat could have gone.  (Examples discussing this question include articles in [The Guardian](http://www.theguardian.com/environment/2015/jun/04/global-warming-hasnt-paused-study-finds), [BBC News](http://www.bbc.com/news/science-environment-28870988), and [Wikipedia](https://en.wikipedia.org/wiki/Global_warming_hiatus)). 

By examining the data here, what evidence do you find or not find for such a pause?  Present an analysis of this data (using the tools & methods we have covered in Foundation course so far) to argue your case.  

One of the evidence is that due to the increasing amount of aerosols in the air, there will be more wave energy reflected back to the universe. Thus, it looks like that there is a "Pause" in global warming. Other proposed evidences include the periodic upwelling of cooler currents in Pacific to absorb heat and the sinking of current in Atlantic to bury remarkable amount of heat into bottom of the sea. However, the new study suggests that the warming rate in the recent 15 years prior to 2014 have just been the same as before. Previous study got the apparent "Pause" result just because it did not account for the change in temperature measuring method across the measuring period.

What additional analyses or data sources would better help you refine your arguments?

Learning about the specific methods of natural systems heat flow modeling and global temperature measurement by the previous studies can provide better insight on the accuracy of the results. Temperature change distribution among different regions around the globe may also be useful. 

## Question 5: Rolling averages
    
- What is the meaning of "5 year average" vs "annual average"?
- Plot the different averages and describe what differences you see and why.  
- Construct 5 year averages from the annual data.  Construct 10 & 20-year averages.
- Plot the different averages and describe what differences you see and why.  

The meaning of "5 year average" is the mean value from a range of five years. However, the annual average is the mean value from just one year. According to the plot from the websites, the 5 year average and 10 year average graphs will be smoother. Due to the climate changes and the global greenhouse effects, we can see the global temperatures are in a constantly increasing trend. This trend will bring many further negative impacts on our environment, and we really need to solve these issues.


# Exercise II: Melting Ice Sheets?

- Data description: <http://climate.nasa.gov/vital-signs/land-ice/>
- Raw data file: <http://climate.nasa.gov/system/internal_resources/details/original/499_GRN_ANT_mass_changes.csv>

## Question 1:

- Describe the data set: what are the columns and units? Where do the numbers come from? 
- What is the uncertainty in measurement? Resolution of the data? Interpretation of missing values?

The columns are data for Time (units: year in decimal), Greenland mass (units: Gross Tonnage), and Antarctica mass (units: Gross Tonnage). The uncertainty behind the measurement is that the ice sheet is losing significant amounts of land-based ice in a very quick speed, the accurate and up-to-date measurement of the land mass is hard to get. The resolution of the data is in meters. The missing values should be the data that cannot be directly acquired by satellites. 


## Question 2:

Construct the necessary R code to import this data set as a tidy `Table` object.

```{r}
icesheet <- read.csv("499_GRN_ANT_mass_changes.csv",
                     col.names=c("Year","Greenland Mass","Antarctica Mass"))

icesheet %>% as_tibble()
icesheet
```

## Question 3:

Plot the data and describe the trends you observe.

```{r}

ggplot(icesheet, aes(x =Year, y =Greenland.Mass)) + geom_line()
ggplot(icesheet, aes(x =Year, y =Antarctica.Mass)) + geom_line()
```

As we can see, the overall trend for both Greenland Mass and Antarctica Mass is downward, and the trend is also winding alone the way. 

# Exercise III: Rising Sea Levels?

- Data description: <http://climate.nasa.gov/vital-signs/sea-level/>
- Raw data file: <http://climate.nasa.gov/system/internal_resources/details/original/121_Global_Sea_Level_Data_File.txt>


## Question 1:

- Describe the data set: what are the columns and units? 
- Where do these data come from? 
- What is the uncertainty in measurement? Resolution of the data? Interpretation of missing values?

The columns are altimeter type, merged file cycle, year+fraction of the year, number of observations, number of weighted observations, GMSL (GIA not applied), Standard Deviation of GMSL (GIA not applied), GMSL (GIA applied), Standard Deviation of GMSL (GIA applied), smoothed GMSL with variation, smoothed GMSL with variation (annual and semi-annual signal removed). These data come from NASA’s Goddard Space Flight Center. The uncertainty in measurement is that the sea level actually is different in different parts of the world. Therefore, it is hard to capture the accurate data that describes the earth's sea level. The resolution of the data is in meters. The missing values are represented as "99900", which are data not captured by the satellites. 

## Question 2:

Construct the necessary R code to import this data set as a tidy `Table` object.
```{r}
globalsea <- read_table("globalsea.txt", 
                  comment="#",
                 col_names = c("altimeter type", "merged file cycle", "year", "number of observations", "number of weighted observations", "GMSL (GIA not applied)", "Standard Deviation of GMSL (GIA not applied)","smoothed GMSL variation (GIA not applied)", "GMSL (GIA applied)", "Standard Deviation of GMSL (GIA applied)", "smoothed GMSL with variation (GIA applied)", "smoothed_GMSL_with_variation_annual_and_semiannual_signal_removed"),
                  na = "99900")
globalsea
```


## Question 3:

Plot the data and describe the trends you observe.

```{r}
ggplot(globalsea, aes(x =year, y =smoothed_GMSL_with_variation_annual_and_semiannual_signal_removed)) + geom_line()
```
As we can see from the graph, the sea level keeps increasing. 

# Exercise IV: Arctic Sea Ice?

- <http://nsidc.org/data/G02135>
- <ftp://sidads.colorado.edu/DATASETS/NOAA/G02135/north/daily/data/N_seaice_extent_daily_v3.0.csv>


## Question 1:

- Describe the data set: what are the columns and units? 
- Where do these data come from? 
- What is the uncertainty in measurement? Resolution of the data? Interpretation of missing values?

The columns are year, month, day, extent, missing values, and source data. The data come from satellite platforms including DMSP, DMSP 5D-3/F17, DMSP 5D-3/F18, Nimbus-7. The uncertainty in measurement is that there are some error measurements in daily data products. The spatial resolution of the data is 25km x 25km, and the temporal resolution of the data is 1 day. The missing values are errors in daily products. 

## Question 2:

Construct the necessary R code to import this data set as a tidy `Table` object.
```{r}
seaice <- read.csv("seaice.csv")
seaice %>% as_tibble()
seaice_agg <- aggregate(seaice[, 4], list(seaice$Year), mean)
seaice_agg
```

## Question 3:

Plot the data and describe the trends you observe.    
```{r}
seaicep <- ggplot(seaice_agg, aes(x =Group.1, y =x)) + geom_line()
seaicep
```


# Exercise V: Longer term trends in CO2 Records


The data we analyzed in the unit introduction included CO2 records dating back only as far as the measurements at the Manua Loa observatory. To put these values into geological perspective requires looking back much farther than humans have been monitoring atmospheric CO2 levels. To do this, we need another approach.


[Ice core data](http://cdiac.ornl.gov/trends/co2/ice_core_co2.html):

Vostok Core, back to 400,000 yrs before present day 

- Description of data set: <http://cdiac.esd.ornl.gov/trends/co2/vostok.html>
- Data source: <http://cdiac.ornl.gov/ftp/trends/co2/vostok.icecore.co2>

## Questions / Tasks: 

- Describe the data set: what are the columns and units? Where do the numbers come from? 
- What is the uncertainty in measurement? Resolution of the data? Interpretation of missing values?
- Read in and prepare data for analysis.
- Reverse the ordering to create a chronological record.  
- Plot data
- Consider various smoothing windowed averages of the data. 
- Join this series to Mauna Loa data
- Plot joined data
- Describe your conclusions

The columns of the data set include depth, age of the ice, mean age of the air, and co2 concentrations. The numbers come from ice samples analyzed by Arctic and Antarctic Research Institute. The uncertainty in measurement is that we cannot capture a whole period of co2 data because the distribution of the ice is not regular. The resolution of the data is in meters. The missing values are data that cannot be extracted by ice core. 

```{r}
icecore <- read_table("icecore.co2.txt", 
                  comment="#",
                 col_names = c("Depth","AgeoftheIce","MeanAgeoftheAir","co2concentrations"))
rev_icecore <- icecore[nrow(icecore):1, ]
icecore
rev_icecore
```

```{r}
icecoreco2 <- ggplot(rev_icecore, aes(x =MeanAgeoftheAir, y =co2concentrations)) + geom_line()
icecoreco2
```
```{r}
co2mauna <- read_table("co2_mauna.txt", 
                  comment="#",
                 col_names = c("Year","Mean","Unc"))
co2mauna
```

```{r}
co2mp <- ggplot(co2mauna, aes(x =Year, y =Mean)) + geom_line()
co2mp
```

As we can see from two graphs, the overall trend of co2 from ancient to now (Vostok Core Data) is winding as co2 increases in a period of time and decreases in a period of time. The trend of co2 concentrations from Mauna Loa data is overall increasing. 
